---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Further topics on random variables

## Transformation of random variables

## Joint Probability Distributions

When events are happened simultaneously, to explore the relationship between two random variables we need **joint probability distributions.**

**Definition**

::: {.callout-important appearance="minimal"}
The function $f(x,y)$ is said to be a **joint density function** of the continuous random variables $X$ and $Y$ if

1.  $f(x,y)\ge0,$ for all $(x,y)$
2.  $\int_{-\infty}^\infty \int_{-\infty}^\infty f(x,y)\ \ dx\ \ dy=1$
3.  $P[(X,Y)\in A]=\int \int _A f(x,y) \ \ dx \ \ dy,$ for any region $A$ in the plane $xy$.
:::

```{r}
library(reticulate)
```

::: panel-tabset
## R

```{r}
#| label: fig-rhist
#| echo: true
set.seed(42)

hist(rnorm(1000),freq = FALSE,ylim = c(0,.4),
     breaks = 10,col = "steelblue",main = "Histogram of Normal distribution")
lines(density(rnorm(1000)),col="blue",lwd=2)

```

## Python

```{python}
#| label: fig-normhist
#| fig-cap: Histogram of Normal Distribution
#| echo: true

import seaborn as sns
import matplotlib.pyplot as plt
import numpy as np
#import pandas as pd

# Generate 1000 samples
seed = 42

n_rv = np.random.normal(loc=0, scale=1, size=1000)
#print(n_rv)


#plt.clf()  # Clears the current figure

## Using `seaborn`

sns.histplot(n_rv, kde=True,stat="density",bins=10)
plt.title("Histogram of Normal Distribution")
plt.legend()
plt.show()
```
:::
